Structuring Theories with Implicit Morphisms

### 2019-02-11 response.txt

Updated response

Below we describe how we worked in all major comments.
We have worked in all minor comments as well.

---------------------------------------------
Comment 1 (Review 1):
  presentation is MMT-specific, unclear what to do with it
  no global vision of what to do with those ingredients and how the proposed concept can help

Comment 2 (Review 1):
  just a bigger example with a dozen theories and information about the advantages provided by implicit morphisms in this setting would be helpful to evaluate its importance

Comment 3 (Review 2) (see also Comment 5):
  the formal theory supporting implicit morphisms is very simple
  only major result of the paper, theorem 1, follows almost immediately, and the paper's contribution reduces to definitions

Response:

It helps to know the following.
Implicit morphism were first conceived by Rabe in 2010 and implemented as part of the Twelf system (which implements the dependent type theory LF).
The theory behind this implementation was never written up and therefore not published.

But the implementation already scaled well, and implicit morphisms were used in the LATIN logic atlas build by Rabe et al. in ~2009-2012.
The LATIN atlas already has around a 1000 theories and atomic morphisms, and about 50 of the latter are marked as implicit.
It also has a few hundred inclusions, each of which induces another implicit morphism.

In the following years, MMT was developed. As part of this development
- implicit morphisms were generalized from LF to the logic-independent level of MMT,
- their theory was worked out and written up,
- they were reimplemented from scratch as a part of MMT.
MMT is backwards-compatible with Twelf, and the LATIN atlas including its implicit morphisms can be used from within MMT.

The present paper was written as the climax of this development to coincide with the first MMT release in which implicit morphisms are officially supported.

The simplicity (Comment 3) is the result of improving syntax, semantics, and implementation for 8 years until we found the simplest possible formulation.
We think of our result as a language design result and of that simplicity as our main achievemnt.
But note that our response to Comment 5 will deepen the results of our paper.

The lack of big picture description (Comment 1) and larger case study (Comment 2) are the result of us focusing on the technical details and keeping the preliminaries and background minimal.
We believe now that was a mistake.

We believe we have remedied all three comments by adding the above text in a new paragraph in the introduction.
We have also increased the size of the example in Sect. 4.2 and 4.3.

---------------------------------------------
Comment 4 (Review 2):
  the hand-waving style led to mistakes or slight inconsistencies (see my comments below)

Response:

We have taken care of inconsistencies and work in all minor comments.
We expect that has naturally eliminated the hand-waving.

---------------------------------------------
Comment 5 (Review 2):
I would have liked to see more discussion on how their theory applies in other settings than MMT, in particular for a diagram of heterogeneous theories (e.g. over a Grothendieck institution).
My guess is that the transfer of results is not that immediate, for example you would need some mechanism for proofs.

Response:

MMT treats logics as theories and logic translations as theory morphisms (which in particular must translate any proof rules delared in the domain logic to proofs in the target logic).
Thus, morphisms between theories and logics are treated uniformly by MMT, and in particular logic morphisms can be implicit in the same way.
Indeed, several logic morphisms (e.g., PL -> FOL -> HOL etc.) are natural candidates for marking as implicit morphisms.

The description of logics and logic morphisms was originally omitted from the paper to make the theory simpler.
In light of Comment 3 and 5, we have added this description to Sect. 2 (which required a few small changes, which can be found by searching for "meta").
We have added a bigger example in Sect. 4.2 and describe in the new Sect. 4.3 how to build it as a diagram of heterogeneous theories.

---------------------------------------------
Noteworthy minor comments from Review 2:

--------
page 6, def. 4:
This seems faulty. Consider the example of Group, without inclusions.
Via flattening, you get
Group?U : type
Group?op : U -> U -> U
because with your definition (n : E)^\flat = t?n : E
I would expect to get
Group?op : Group?U -> Group?U -> Group?U
This must also be taken care of for imports.

Response:
The presentation in the paper is slightly confusing because "U" is concrete syntax, which is parsed in "Group?U" in abstract syntax.
We have added Rem. 1 to clarify that.

--------
page 10, renamings:
when using the attribute renaming, are the conditions of the
definition checked?

Response:
Yes.

--------
page 12:
instead of introducing "implicit-later", would it not be better to
have a mechanism for deriving that the morphism is implicit
(e.g. because the isomorphism proof exists)?

Response:
In general, not every isomorphism can automatically be implicit because there may be multiple different isomorphisms between the same theories.
However, in this particular case, the system could indeed infer that G2DG is the inverse of an implicit isomorphism and therefore should be implicit as well.
We rephrased the example accordingly.

---------------------------------------------
Comment 6 (Review 3):
  not convinced of the practical benefits of this particular implementation, and tend to consider that most of them can also be achieved (as per Section 4 of the paper), with minimal or no additional effort, using ordinary theory morphisms
  would gladly accept a counterargument

Response:
We agree that the effect can be achieved with minimal additional effort.
But we hold that implicit morphisms, which are ordinary theory morphisms with an additional boolean flag, are already the simplest possible way to do it, and we've spent years finding whittling down the ideas to the simplest possible solution.
(See also our response to Comment 1-3.)

This comment had 3 sub-comments, to which we respond below.
We believe that these responses constitute a strong counterargument.

------------
Comment 6.1: In regard to 'Identifying theories via implicit isomorphisms'

Response:
The reviewer is right with one exception: The main advantage of implicit morphisms m:A -> B is that m is later on inserted automatically without the user having to refer to m every time.
That is only possible if the user manually clarifies which morphisms should be implicit because there could be more than one isomorphism A -> B.

See also the response to Comment 6.2.

------------
Comment 6.2: In regard to 'Fine-granular and flexible theory hierarchies'

Response:
The well-formedness property of implicit morphisms is the same as for ordinary morphisms.
In particular, nothing special happens when the user makes the morphism CancellationMonoid -> Group implicit (except for checking that there is no implicit morphism of that type yet).

All the advantage come later when Group is used:

* A later theory that includes Group can directly refer to all identifiers from CancellationMonoid.
  MMT even lists these identifiers automatically in the auto-completion dialog, and the MMT parser automatically uses their notations.
  Thus, users can simply include Group as if it includes CancellationMonoid without having to know the details of the library strucutre.

* Any parameter of type CancellationMonoid can be substituted with an instance of Group.
  Similarly, any morphism into CancellationMonoid can be composed with a morphism out of Group. 
  In both cases, MMT automatically inserts the implicit moprhism.

Overall, this yields a big advantage when releasing a large library because users can work more intuitively without having to learn the library structure in detail (e.g., a user can use Group like a CancellationMonoid without worrying about the conversion).

We have added a paragraph at the end of Sect. 3.1 to emphasize these advantages.
Moreover, we have revised 4.2 to strengthen the motivation and use a larger example.

Regarding the question about inclusions:
Because inclusions are implicit by default, the commutativity requirement already ensures that an implicit morphism A -> B must be the identity on all theories that are included into both A and B.

------------
Comment 6.3: In regard to 'Transparent refactoring'

Response:
The reviewer is right.
A smarter way to do the refactoring example is to leave s as it is and introduce a second implicit morphism s' -> s.
We have changed the example accordingly.

The reviewer is also right that external refactoring mechanisms should also be used.
But when managing releases of big libraries, it is often desirable to ensure backwards compatibility without requiring users to run a refactoring propragation tool.
A typical application would be to deprecate the original theories but use implicit morphisms to ensure that they still work.
Then after a transition period the original theories are removed from the library and users are expected to switch to the new, cleaner and more orderly diagram.

We have added some sentences on that.

---------------------------------------------
Comment 7 (Review 3):
  discussion on the potential uses of implicit morphisms outside the scope of MMT
  connection with other generalisations of the set-theoretic notion of inclusion, e.g., factorisation or inclusion systems).

Response:
We have no results in this direction yet.
We thought of factorization and inclusion systems but did not mention them because we had no results about how they relate to implicit morphisms.
We'd be happy to receive ideas on this issue.
We added a sentence on that under future work.

### 2019-02-11 appl.tex

A common need in developing large libraries (both in formal and informal developments) is to identify two theories $S$ and $T$ via a canonical choice of isomorphisms.
In these cases, it is often desirable to use $S$-syntax and $T$-syntax interchangeably.
But one of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
In this section, we show that implicit morphisms elegantly allow exactly that kind of intuitive identification.

In the sequel, we present several ways to obtain implicit isomorphisms conveniently.
In general, note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
Moreover, making a pair of isomorphisms implicit is well-formed if there are no other implicit morphisms between $S$ and $T$ yet.

@@

\begin{example}[Renaming]
Consider a variation of the theory $\cn{Monoid}$ from Ex.~\ref{syn:incl} in a different library:
\begin{mmtcode}
Monoid2 =
  M          : type
  connective : M $\to$ M $\to$ M	 # 1 $\circ$ 2 
  neutral    : M
\end{mmtcode}
This theory is isomorphic to the previously introduced theory \cn{Monoid} under the trivial renaming
	\begin{mmtcode}
renaming MonoidRen : Monoid2 -> Monoid =
  M          := U
  connective := op 
  neutral    := unit
\end{mmtcode}
\end{example}

@@

\begin{example}[Extension with a Theorem]
We extend the theory $\cn{Group}$ from Ex.~\ref{syn:incl} with a theorem
\begin{mmtcode}
InverseInvolution =
  definitional include Group
  inverse_invol : $\forall$[x] (x$^{-1}$)$^{-1}$$\doteq$ x  =  (proof omitted)
\end{mmtcode}
\end{example}

@@

\begin{example}\label{group:iso}
Having declared the morphism $\cn{DG2G}$ (as in Ex.~\ref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
\begin{mmtcode}
implicit G2DG : Group -> DivGroup =
  U       := U
  op      := [a,b] a/(unit/b) 
  unit    := unit
  inverse := [a] unit/a
\end{mmtcode}
\end{example}

While making one of these isomorphisms implicit is straightforward, doing both requires checking that $m$ and $n$ are actually isomorphisms.
Otherwise, the uniqueness condition would be violated.
Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $(c:E)\in\flt{A}$.
Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.

@@

\begin{example}[Isomorphisms]
With the above extension in place, we can make both isomorphisms $m$ and $n$ from above implicit:
 \[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
 \[\keyword{implicit-later}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
 \[\keyword{equal}\cn{G2DG};\cn{DG2G}=\id{\cn{Group}}:\cn{Group}\to\cn{Group}=\text{(omitted)}\]
 \[\keyword{equal}\cn{DG2D};\cn{G2DG}=\id{\cn{DivGroup}}:\cn{DivGroup}\to\cn{DivGroup}=\text{(omitted)}\]
where the isomorphisms are as above and we omit all the equality proofs.

Note that we cannot mark \cn{G2DG} as \keyword{implicit} because that is only well-formed after proving the equalities.
So we use the new attribute \keyword{implicit-later}, which states that a morphism should be considered as implicit as soon as subsequent equality proofs make it well-formed to do so.
\end{example}

\subsection{Fine-Granular and Flexible Theory Hierarchies}
\input{example}

@@

A major drawback of using modular theories is that it can preclude transparent refactoring.
For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $s$ to $r$.
This change should be straightforward as it does not change the semantics of $t$.

However, this is not a local change.
It also requires updating every qualified reference from $s?n$ to $r?n$.
Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know or does not have access to.

With implicit morphisms, we can solve this problem by making only the following local changes:
\begin{compactenum}
  \item We rename $s$ to $s'$.
  \item We delete the declaration $D$ from $s'$.
  \item We create new theories $r'=\{\icl{r},\,D\}$ and $s=\{\icl{s'},D\}$.
  \item We change $t$ to $t=\{\icl{r'},\icl{s'}\}$.
  \item We add an implicit morphism $s\to t$ that maps $s?n$ to $r'?n$.
\end{compactenum}
The situation before (left) and after (right) refactoring is given below:
\begin{center}
\begin{tikzpicture}[scale=1]
  \node (t) at (0,0) {$t$};
  \node (r) at (-1,-2) {$r$};
  \node (s) at (1,-1) {$s$};
  \draw[right hook-latex] (r) to (t) {};
  \draw[right hook-latex] (s) to (t) {};

@@

  \node (r) at (-1,-2) {$r$};
  \node (sp) at (1,-2) {$s'$};
  \draw[right hook-latex] (r) to (rp) {};
  \draw[right hook-latex] (sp) to (s) {};

@@

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\end{center}
Afterwards, $t$ has the desired new structure.
But all old references to $s?n$ stay well-formed so that no global changes are needed.

>>>>

A common need in developing large libraries (both in formal and informal developments) is to identify two theories $S$ and $T$ via a canonical choice of isomorphisms.
In these cases, it is often desirable to use $S$-syntax and $T$-syntax interchangeably.
But one of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
In this section, we show that implicit morphisms elegantly allow exactly that kind of intuitive identification: we mark the canonical isomorphisms as implicit.

In the sequel, we present several ways to obtain implicit isomorphisms conveniently.
In general, note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
Moreover, making a pair of isomorphisms implicit is only allowed if there are no other implicit morphisms between $S$ and $T$ yet.

@@

\begin{example}[Renaming]
Consider a variation of the theory $\cn{Monoid}$ from Ex.~\ref{syn:incl} in a different library:
\begin{mmtcode}
Monoid2: $\Log$ =
  M          : type
  connective : M $\to$ M $\to$ M	 # 1 $\circ$ 2 
  neutral    : M
\end{mmtcode}
This theory is isomorphic to the previously introduced theory \cn{Monoid} under the trivial renaming
	\begin{mmtcode}
renaming MonoidRen : Monoid2 -> Monoid =$^{\id{\Log}}$
  M          := U
  connective := op 
  neutral    := unit
\end{mmtcode}
\end{example}

@@

\begin{example}[Extension with a Theorem]
We extend the theory $\cn{Group}$ from Ex.~\ref{syn:incl} with a theorem
\begin{mmtcode}
InverseInvolution: $\Log$ =
  definitional include Group
  inverse_invol : $\forall$[x] (x$^{-1}$)$^{-1}$$\doteq$ x  =  (proof omitted)
\end{mmtcode}
\end{example}

@@

\begin{example}\label{group:iso}
Having declared the morphism $\cn{DG2G}$ (as in Ex.~\ref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
\begin{mmtcode}
implicit G2DG : Group -> DivGroup =$^{\id{\Log}}$
  U       := U
  op      := [a,b] a/(unit/b) 
  unit    := unit
  inverse := [a] unit/a
\end{mmtcode}
\end{example}

While making one of these isomorphisms implicit is straightforward, doing it for both requires checking that $m$ and $n$ are actually isomorphisms.
Otherwise, the uniqueness condition would be violated.
Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $(c:E)\in\flt{A}$.
Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.

@@

\begin{example}[Isomorphisms]
With the above extension in place, we can make both isomorphisms $m$ and $n$ from above implicit:
\[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
\[\keyword{implicit}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
\[\keyword{equal}\cn{G2DG};\cn{DG2G}=\id{\cn{Group}}:\cn{Group}\to\cn{Group}=\text{(omitted)}\]
\[\keyword{equal}\cn{DG2D};\cn{G2DG}=\id{\cn{DivGroup}}:\cn{DivGroup}\to\cn{DivGroup}=\text{(omitted)}\]
where the isomorphisms are as above and we omit all the equality proofs.

There is a subtle difficulty in marking \cn{G2DG} as \keyword{implicit}: the implicit-diagram only commutes after proving the equalities, which are only declared later.
One option is to delay the commutativity check until the equalities have been checked.
In that case, care must be taken to avoid using the implicitness of \cn{G2DG} while proving the equalities.
But we obtain a more elegant solution from the observation that it is always sound and harmless to automatically make the inverse of an implicit isomorphism implicit as well.
Thus, we can omit the attribute \keyword{implicit} on \cn{G2DG} altogether and use an implementation that infers that \cn{G2DG} is the inverse of an implicit isomorphism.
\end{example}

\subsection{Fine-Granular and Flexible Theory Hierarchies}\label{sec:granular}
\input{example}

\subsection{Logic Morphisms}

So far, we have assumed all theories use the same fixed meta-theory, e.g., $\Log$ in the running examples.
But in practice, we usually develop theories heterogeneously by using the weakest possible meta-theory for each module.
A strength of \mmt is that meta-theories are normal theories so that the same structuring formalism can be used for them, e.g., morphisms, includes, pushouts, and implicit morphisms are directly available for meta-theories.

For example, consider the example from Figure~\ref{fig:magmas}.
We could use first-order logic \cn{FOL} as the common meta-theory of all theories.
But actually the much weaker logic \cn{Horn} (which uses restricted logical connectives that only allow creating Horn formulas) is sufficient for most of them.

We do not want to declare a direct include $\cn{Horn}\to\cn{FOL}$ for the same reasons as discussed in  Section~\ref{sec:granular}.
Instead, we want to give a morphism $\cn{EmbedHorn}:\cn{Horn}\to\cn{FOL}$, which maps all Horn formula constructors to their \cn{FOL} counterparts.
We can now make \cn{EmbedHorn} implicit, thus capturing the fact that Horn logic is a fragment of \cn{FOL}.

Moreover, assume we have built the diagram in Figure~\ref{fig:magmas} using the meta-theory \cn{Horn} where possible and \cn{FOL} where necessary.
An example for the latter is \cn{TotalOrder}, which also uses $\vee$.
Without implicit morphisms, we would have to write
\[\cn{TotalOrder}:\cn{FOL}=\{\icl{\cn{EmbedHorn}(\cn{PartialOrder})},\ldots\}\]
Here the \cn{Horn}-theory \cn{PartialOrder} must be explicitly translated to \cn{FOL} before including it, which is awkward for users both when writing and reading.
But using implicit morphisms and the semantics of Ex.~\ref{impl:incl2}, we can simply write $\icl{\cn{PartialOrder}}$ --- the logic translation remains transparent to the user.

In practice, many logic morphisms are naturally implicit, either because they are includes to begin with or because they represent a canonical logic embedding.

@@

A major drawback of using modular theories is that it can preclude transparent refactoring (to insert intermediate theories as mentioned in Section~\ref{sec:granular}).
As an example, we consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $s$ to $r$.
This change should be straightforward as it does not change the semantics of $t$.

However, this is not a local change.
It also requires updating every qualified reference from $s?n$ to $r?n$.
Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know about or does not have access to.
One option in this case is to use an extra-logical refactoring tool that propagates such changes.
But when managing releases of big libraries, it is often desirable to deprecate the original theories but still ensure backwards compatibility.
Then after a transition period the original theories are removed from the library and users are expected to propagate the refactoring.

With implicit morphisms, we can solve this problem by making only the following local changes:
\begin{compactenum}
  \item We keep $r$ and $s$ as they are.
  \item We create a new theory $r'=\{\icl{r},\,D\}$.
  \item We create a new theory $s'$ that is like $s$ except for deleting the declaration $D$.
  \item We change $t$ to $t=\{\icl{r'},\icl{s'}\}$.
  \item We add implicit morphisms $s'\to s$ (mapping $s'?x$ to $s?x$ for all $x$) and $s\to t$ (mapping $s?n$ to $r'?n$ and $s?x$ to $s'?x$ for all $x\neq n$).
\end{compactenum}
The situation before (left) and after (right) refactoring is given below.
Note that the right diagram commutes.
\begin{center}
\begin{tikzpicture}[scale=1]
  \node (t) at (0,0) {$t$};
  \node (r) at (-1,-2) {$r$};
  \node (s) at (1,-1) {$s$};
  \draw[right hook-latex] (r) to (t) {};
  \draw[right hook-latex] (s) to (t) {};

@@

  \node (r) at (-1,-2) {$r$};
  \node (sp) at (1,-2) {$s'$};
  \draw[right hook-latex] (r) to (rp) {};
  \draw[arrow] (sp) to (s) {};

@@

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\end{center}
Afterwards, $t$ has the desired new structure.
But all old references to $r$ and $s$ stay well-formed so that no global changes are needed.
Now $r$ and $s$ can be deprecated and eventually removed in favor of $r'$ and $s'$.

### 2019-02-11 example.tex

A common problem when defining modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
Instead, we could include \cn{Monoid} into the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
This change is not easy in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents a very expensive refactoring problem.
So instead we could systematically build a hierarchy that uses every intermediate theory (as done in \cite{mathscheme}).
But this yields a very deep and complex hierarchy that is hard to navigate for casual users.
Moreover, it does not protect us from later on discovering yet another intermediate theory that should have been added.

Implicit morphisms provide a simple solution to this problem because they behave effectively like inclusions but can be added later on.
In the above example, we would
\begin{compactitem}
 \item define \cn{CancellationMonoid} with an include from \cn{Monoid},
 \item keep \cn{Group} as it is, i.e., also with an include \cn{Monoid},
 \item add an implicit morphism $\cn{CancellationMonoid}\to\cn{Group}$.
\end{compactitem}
\medskip

As a more complex case study, we have used our implementation of implicit morphisms in \mmt to formalize a hierarchy of theories between \cn{Band} and \cn{Semilattice}:% 
\footnote{Our formalization can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
\begin{center}
\begin{tikzpicture}[xscale=2]
  \node (s1) at (0,0) {$z\circ x\circ y\circ z\doteq  z\circ x\circ z\circ y\circ z$};

  \node (s2a) at (-1,-1) {$z\circ x\circ y\doteq  z\circ x\circ z\circ y$};
  \node (s2b) at (1,-1) {$y\circ x\circ z\doteq  y\circ z\circ x\circ z$};
  \draw[arrow] (s1) to (s2a) {};
  \draw[arrow] (s1) to (s2b) {};

  \node (s3a) at (-2,-2) {$x\circ y\doteq  x\circ y\circ x$};
  \node (s3b) at (0,-2) {$z\circ x\circ y\circ z\doteq  z\circ y\circ x\circ z$};
  \node (s3c) at (2,-2) {$x\circ y\doteq  y\circ x\circ y$};
  \draw[arrow] (s2a) to (s3a) {};
  \draw[arrow] (s2a) to (s3b) {};
  \draw[arrow] (s2b) to (s3b) {};
  \draw[arrow] (s2b) to (s3c) {};

  \node (s4a) at (-1,-3) {$z\circ x\circ y\doteq  z\circ y\circ x$};
  \node (s4b) at (1,-3) {$x\circ y\circ z\doteq  y\circ x\circ z$};
  \draw[arrow] (s3a) to (s4a) {};
  \draw[arrow] (s3b) to (s4a) {};
  \draw[arrow] (s3b) to (s4b) {};
  \draw[arrow] (s3c) to (s4b) {};

  \node (s5) at (0,-4) {$x\circ y\doteq  y\circ x$};
  \draw[arrow] (s4a) to (s5) {};
  \draw[arrow] (s4b) to (s5) {};

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\end{center}
All of these theories are of the form $t=\{\icl{\cn{Band}},\,a: F\}$, where $F$ is the formula given in the diagram above.%
%\footnote{All varieties of bands can be axiomatized in this way.}
In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a: \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.

The diagram above gives all morphisms between them that describe the lattice structure of the corresponding varieties of bands, all of which map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
In our formalization, we make all of these morphisms implicit.
It is straightforward to prove that the diagram commutes: any two morphisms are identical except for the assignment to the axiom $a$, and these are equal due to proof irrelevance.

>>>>

A common problem when defining modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
Instead, we could have used an intermediate theory and includes $\cn{Monoid}\harr \cn{CancellationMonoid}\harr\cn{Group}$.

It is very common to have increasingly strong theories $R,S,T$, where a design with two includes $R\harr S\harr T$ is not desirable:
\begin{compactitem}
 \item Often $R\harr T$ has been defined first and $S$ only later.
   This is very common because people usually formalize the most important theories (e.g., \cn{Monoid} and \cn{Group}) first.
   But inserting $S$ is not easy in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents a very expensive refactoring problem.
   And even if we systematically use includes for every known intermediate theory like $S$ (as done in \cite{mathscheme}), we might later discover a new intermediate theory that should have been added.
 \item Often the most natural axioms to use in $T$ are the same independent of whether $T$ includes $R$ or $S$ (e.g., users might prefer the usual inverse-element axioms in \cn{Group} even if they have included \cn{CancellationMonoid}).
 In that case, the axioms of $S$ become provable in $T$ if we use $R\harr S\harr T$.
 This either causes $T$ to have redundant axioms or requires a more complex include mechanism that allows $T$ to include $S$ in a way that turns some of $S$-axioms into theorems.
\end{compactitem}

Therefore, it is common to use a commuting triangle consisting of two includes $R\harr T$ and $R\harr S$ and one morphism $m:S\to T$.
But this is awkward because the relation ``every $T$ is an $S$'' is now mediated by $m$ rather than being canonical.  
Implicit morphisms provide a simple solution to this problem: we keep the triangle but make the morphism $m$ implicit.
This captures exactly the canonical conversion from $S$ to $T$.

\begin{figure}[htb]
	\begin{center}
		\includegraphics[width=.8\textwidth]{graph.png}
		\caption{Magma hierarchy with includes (gray) and implicit morphisms (black)}\label{fig:magmas}
	\end{center}
\end{figure}

Already the elementary algebraic hierarchy provides countless examples of such situations.
For a small fragment of the hierarchy of magmas, Figure~\ref{fig:magmas} shows one possible design using numerous implicit morphisms.
In particular, it uses some of the examples and features from this paper, e.g., an implicit isomorphism to identify the order-theoretic and the algebraic development of semilattices.

It also uses multiple implicit morphisms to introduce the various intermediate theories between $\cn{Band}\harr\cn{SemiLattice}$. 
All of these are of the form $t=\{\icl{\cn{Band}},\,a: F\}$, e.g., \cn{LeftRegularBand} uses $F=\forall x,z. z\circ x\circ z \doteq z\circ x$.
The implicit morphisms map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
It is straightforward to prove that this part of the diagram commutes: any two morphisms are identical except for the assignment to the axiom $a$, and these are equal due to proof irrelevance.%
\footnote{Our formalization of bands can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
%\footnote{All varieties of bands can be axiomatized in this way.}

%\begin{center}
%\begin{tikzpicture}[xscale=2]
%  \node (s1) at (0,0) {$z\circ x\circ y\circ z\doteq  z\circ x\circ z\circ y\circ z$};
%  
%  \node (s2a) at (-1,-1) {$z\circ x\circ y\doteq  z\circ x\circ z\circ y$};
%  \node (s2b) at (1,-1) {$y\circ x\circ z\doteq  y\circ z\circ x\circ z$};
%  \draw[arrow] (s1) to (s2a) {};
%  \draw[arrow] (s1) to (s2b) {};
%  
%  \node (s3a) at (-2,-2) {$x\circ y\doteq  x\circ y\circ x$};
%  \node (s3b) at (0,-2) {$z\circ x\circ y\circ z\doteq  z\circ y\circ x\circ z$};
%  \node (s3c) at (2,-2) {$x\circ y\doteq  y\circ x\circ y$};
%  \draw[arrow] (s2a) to (s3a) {};
%  \draw[arrow] (s2a) to (s3b) {};
%  \draw[arrow] (s2b) to (s3b) {};
%  \draw[arrow] (s2b) to (s3c) {};
%  
%  \node (s4a) at (-1,-3) {$z\circ x\circ y\doteq  z\circ y\circ x$};
%  \node (s4b) at (1,-3) {$x\circ y\circ z\doteq  y\circ x\circ z$};
%  \draw[arrow] (s3a) to (s4a) {};
%  \draw[arrow] (s3b) to (s4a) {};
%  \draw[arrow] (s3b) to (s4b) {};
%  \draw[arrow] (s3c) to (s4b) {};
%
%  \node (s5) at (0,-4) {$x\circ y\doteq  y\circ x$};
%  \draw[arrow] (s4a) to (s5) {};
%  \draw[arrow] (s4b) to (s5) {};
% 
%  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
%  %\node at (t1.north west) {$S$};
%  %\node at (t1) {$D$};
%  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
%  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
%  %\node at (t2.north west) {$T$};
%  %\node at (t2) {$C$};
%  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
%\end{tikzpicture}
%\end{center}

### 2019-02-11 implicit.tex

@@

\begin{definition}[Obtaining Implicit Morphisms]
We define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
\begin{itemize}
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
\begin{nomodexp}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
\end{nomodexp}
 \item all morphisms that additional language features designate as implicit based on the use of additional attributes.
\end{itemize}

@@

>>>>

In practice, the \mmt system searches for an implicit morphism $\ipc{i}{S}{T}$ whenever needed to make an expression well-formed, e.g.,
\begin{compactitem}
 \item An $S$-constant $c$ is used in $T$: treat $c$ as an abbreviation for $i(c)$.
 \item A model $M$ of $T$ is used even though a model of $S$ is expected: reduce $M$ via $i$.
 \item A morphism $M:R\to S$ is composed with a morphism $M':T\to U$: treat $M;M'$ like $M;i;M'$.
\end{compactitem}

@@

\begin{definition}[Obtaining Implicit Morphisms]
We define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
\begin{itemize}
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
 \item if $t$ has meta-theory $L$, the identity map as an implicit morphism $L\to t$,
\begin{nomodexp}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
\end{nomodexp}
 \item all morphisms that additional language features designate as implicit based on the use of additional attributes.
\end{itemize}

@@

With implicit morphisms, we can also relax the semantics of many structuring features:

\begin{example}[Includes and Meta-Theories (continued from Ex.~\ref{sem:incl})]\label{impl:incl2}
Consider $t:L=\{\ldots,\icl{S},\ldots\}$ where $S$ has meta-theory $L'$.
Instead of requiring $L'\harr L$ as in Ex.~\ref{sem:incl}, we require only $\ipc{i}{L'}{L}$.

In that case, we treat $\icl{S}$ as an abbreviation for $\icl{i(S)}$, where $i(S)$ is the pushout of $S$ along $i$ (see \cite{rabe:howto} for details on \mmt pushouts).
If $L'\harr L$, this reduces to the original semantics.
\end{example}
